Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2023-12-23 demystify.tex
Updates from Overleaf

\begin{enumerate}
  \item Should you apply to a CS PhD in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically 3 faculty (\S\ref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
        \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
          \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
          \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
          \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
          \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
          \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
          \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
        \end{itemize}
  \item Miscs and FAQS
        \begin{itemize}
          \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
          \item Your PhD costs about \$400K in total, but you do not pay for it (\S\ref{sec:ra-cost}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive labmates, healthy working environment ... and are happy (\S\ref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

Second, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS (or having to know a topic/proposal/adviser in advance). If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \S\ref{sec:non-us-differences} compares CS PhD study in the US to other countries.

@@

In many cases you do not have to do a PhD in CS to work in your area of interested. For example, in addition to a traditional CS department, GMU has IST and Cybersecurity departments, which have faculty  with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is totally possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is  common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in CS department.

However, if your intention is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. In fact, if a faculty is not in CS, it is unlikely that they can be the \emph{main} adviser of a CS PhD student---they may \emph{co-advise} or be in the PhD dissertation committee, but your main adviser will need to be a tenured or tenure-track faculty in CS. If in doubt, you should check with the CS department for their requirements.  For this specific reason,  CSRankings includes only tenured or tenured track faculty who can advise CS PhD students. I also have compiled a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{list} of Vietnamese faculty who can advise PhD students.

@@

\emph{Project proposal}: in many countries, you have to choose a project and adviser \emph{during} the application process (e.g., you write a proposal to a potential adviser). In the US, you often start your PhD without an adviser or project and find them later. This is also why US PhD programs are longer.

@@

\emph{Faculty position after PhD}: In other countries, PhD graduates who are interested academia typically apply for research positions at research labs, i.e., postdocs, and then consider faculty position. In the US, PhD graduates often apply directly for faculty position (postdoc for US graduates is no longer a popular option as it was before).  This is why US PhD programs are longer and more focused on research.

@@

This section lists various cultural issues that international students might want to pay attention to.

@@

There's a misconception that in the US it's all business, with professors as bosses who pay students for their work, and you doing something nice implying you expect something in return. However, the reality is quite the opposite. While people can be straightforward and direct, they are also friendly and informal. You can call your professor by their first name (\S\ref{sec:address}), disagree with them and argue (and gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks.  Many students and professors maintain lifelong relationships, staying in touch through cards, emails, and calls, even after their academic journey ends.% These connections are genuine and meaningful.

>>>>

\begin{enumerate}
  \item Should you apply to a CS PhD in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically three faculty (\S\ref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
        \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
          \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
          \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
          \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
          \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
          \item Getting an interview is typically a \emph{good sign} (no one interviews weak candidates).
        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TA and RA are two main sources of funding.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} be adviser, i.e., cannot graduate PhD students  (\S\ref{sec:schoolsandprofs}).
          \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
        \end{itemize}
  \item Miscs and FAQS
        \begin{itemize}
          \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
          \item Your PhD costs about \$400K in total, but you \emph{do not} pay for it (\S\ref{sec:ra-cost}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\S\ref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

Second, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or knowing a topic/proposal/adviser in advance (by the way, GRE is also not required!). If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \S\ref{sec:non-us-differences} compares CS PhD study in the US to other countries.

@@

In many cases you do not need to do a PhD in CS to work in your area of interested. For example, in addition to a traditional CS department, GMU has IST and Cybersecurity departments, which have faculty  with PhD in CS and work on CS topics (e.g., AI, Security, Robotics).  So it is totally possible that you still get to do CS research and publish in CS-focused venues even if you're not in a traditional CS program.  It is  common to see faculty with CS PhD in a non-CS department as well as faculty with non-CS PhD in CS department.

However, if your goal is a PhD in CS, then you likely need to be in the CS dept \emph{and} advised by a CS faculty. In fact, if a faculty is not in CS, it is unlikely that they can be the \emph{main} adviser of a CS PhD student. They may \emph{co-advise} or be in the PhD dissertation committee, but your main adviser will need to be a tenured or tenure-track faculty in CS. If in doubt, you should check with the CS department for their requirements.  For this specific reason,  CSRankings includes only tenured or tenured track faculty who can advise CS PhD students. I also have compiled a \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{list} of Vietnamese faculty who can advise PhD students.

@@

\emph{Project proposal}: in many countries, you have to choose a project and adviser \emph{during} the application process (e.g., you write a proposal to a potential adviser). In the US, you often start your PhD without an adviser or project and find them later. This is because as someone who just finished your undergraduate study, it is unlikely that you know enough about research to do proposal or picking an adviser.  This is also why US PhD programs are longer.

@@

\emph{Faculty position after PhD}: In other countries, PhD graduates interested academia typically apply for additional research appointments, i.e., postdocs, and then consider faculty position. In the US, PhD graduates often apply directly for faculty position (postdoc for US graduates is no longer a popular option as it was before).  %This is also why US PhD programs are longer (remember, 5--7 years).

@@

This section lists some cultural issues that international students might want to pay attention to. These issues are not specific to just the PhD admission process but in general.

@@

\section{Relationships with your Profs.}
There's a misconception that in the US it's all business, with professors as bosses who pay students for their work, and you doing something nice implying you expect something in return. However, the reality is quite the opposite. While people can be straightforward and direct, they are also friendly and informal. You can call your professor by their first name (\S\ref{sec:address}), disagree with them and argue (and gain respect doing so), seek their help (even on personal matters), come to their houses for parties or gathering, and give them small thoughtful gifts that they put on their desks.  Many students and professors maintain lifelong relationships, staying in touch through cards, emails, and calls, even after their academic journey ends.% These connections are genuine and meaningful.
